---
title: "Effects on Learning of Small Class Sizes"
author: "Majoni Maumbe"
date: June 3rd, 2018 
output:
  word_document: default
  html_document:
    df_print: paged
---

#Data Sources and Packages
```{r}
#Dataset: https://vincentarelbundock.github.io/Rdatasets/csv/Ecdat/Star.csv
#Data Dictionary: https://vincentarelbundock.github.io/Rdatasets/doc/Ecdat/Star.html

#Packages
# install.packages("ggplot2")
# install.packages("reshape2")
# install.packages("plotly")
# install.packages("plyr")
# install.packages("dplyr")
# install.packages("arules")
# install.packages("arulesViz")
# install.packages("rpart")
# install.packages('rattle')
# install.packages('rpart.plot')
# install.packages('RColorBrewer')
# install.packages("Cairo")
# install.packages("CORElearn")
# install.packages("caret")
# install.packages("e1071")
# install.packages("knitr")
# install.packages("cluster")
# install.packages("factoextra")

suppressMessages(library(ggplot2))
suppressMessages(library(reshape2))
suppressMessages(library(plotly))
suppressMessages(library(plyr))
suppressMessages(library(dplyr))
suppressMessages(library(arules))
suppressMessages(library(arulesViz))
suppressMessages(library(rpart))
suppressMessages(library(rattle))
suppressMessages(library(rpart.plot))
suppressMessages(library(RColorBrewer))
suppressMessages(library(Cairo))
suppressMessages(library(CORElearn))
suppressMessages(library(caret))
suppressMessages(library(e1071))
suppressMessages(library(knitr))
suppressMessages(library(cluster))
suppressMessages(library(factoextra))
```

#Loading and Cleaning the Dataset
```{r}
#Load the dataset and remove ID column
CSdata <- read.csv("https://vincentarelbundock.github.io/Rdatasets/csv/Ecdat/Star.csv")
CSdata <- CSdata[,-1]

#Rename the column headers
colnames(CSdata) <- c("math_score", "reading_score", "class_size", "teacher_experience", "student_gender", "free_lunch", "race", "school_id")

#Append Total Score Column and Re-Order Columns
CSdata$total_score <- CSdata$math_score+CSdata$reading_score
CSdata <- CSdata[,c(8,3,5,6,7,4,1,2,9)]
```


####Check for Missing Values
```{r}
colSums(is.na(CSdata))
```


####Data Dictionary
```{r}
field_name <- colnames(CSdata)
field_discriptions <- c("school indicator variable", 
                        "type of class, a factor with levels (regular,small.class,regular.with.aide)",
                        "a factor with levels (boy,girl)",
                        "qualified for free lunch ?",
                        "a factor with levels (white,black,other)",
                        "years of total teaching experience",
                        "total math scaled score",
                        "total reading scaled score",
                        "total scaled score")
dictionary <- data.frame("Field Name" = field_name,"Field Discriptions" = field_discriptions)
kable(dictionary)
```


####Cleaned Dataset Preview
```{r}
#Print results of first 10 records
kable(head(CSdata, 5))
```


####Formatting nominal values to factors with levels
```{r}
CSdata$class_size <- factor(CSdata$class_size, levels = c("small.class", "regular", "regular.with.aide"))
CSdata$student_gender <- factor(CSdata$student_gender, levels = c("boy", "girl"))
CSdata$free_lunch <- factor(CSdata$free_lunch, levels = c("yes", "no"))
CSdata$race <- factor(CSdata$race, levels = c("white", "black", "other"))

#View Structure of the Dataframe
str(CSdata)
```

#Summary Statistics of the Dataframe

####Integer Value Statistics
```{r}
int_stat <-  summary(CSdata[,c(6,7,8,9)])
colnames(int_stat) <- c("Teacher Experience", "Math Score", "Reading Score", "Total Score")
kable(int_stat)

```

####Nominal Value Statistics
```{r}
nom_stat <-  summary(CSdata[,c(2,3,4,5)])
colnames(nom_stat) <- c("Class Size", "Student Gender", "Free Lunch", "Race")
kable(nom_stat)
```

####Average Math and Reading Scores by Class Size
```{r}
csdatamean <- aggregate(CSdata[,7:9], by=list(CSdata$class_size), FUN=mean)
colnames(csdatamean) <- c("Class Size", "Math Score", "Reading Score", "Total Score")
kable(csdatamean)
```


#Visualizing the Data

####Does Teaching Experience and Class Size Affect Total Score?
```{r, fig.width=8}

agg <- aggregate(CSdata$total_score, by=list(CSdata$class_size,CSdata$teacher_experience), 
  FUN=mean, na.rm=TRUE)
names(agg) <- c("class_size", "teacher_experience", "total_score")

ggplot(agg, aes(x=teacher_experience, y=total_score, group=class_size, colour=class_size ) ) + 
            geom_line(size=2) + labs(x="Teacher Experience (years)",y="Total Score") +
            ggtitle("Smaller class sizes appear to have a higher average total score") + 
            theme(legend.position="bottom")
```


####Does a student's ethnicity play a factor?
```{r, fig.width=8}
ggplot(CSdata, aes(x=race, y=total_score, color=class_size)) +
  geom_boxplot() +
  labs(x="Student Ethnicity",y="Total Score") +
            ggtitle("The median total scores are higher for student in smaller classes and higher\namong white students") + 
  theme(legend.position="bottom")


```


#Transformation of Test Scores to Z-scores

#####Calculating the Test Score's Z-Score by Class Size
```{r}
#Z-Score of Small Classes Math and Reading
CSZ_score <- CSdata
CSZ_score$z_score_math <- scale(CSdata$math_score, center = TRUE, scale = TRUE)
CSZ_score$z_score_reading <- scale(CSdata$reading_score, center = TRUE, scale = TRUE)
CSZ_score$z_score_total <- scale(CSdata$total_score, center = TRUE, scale = TRUE)

kable(head(CSZ_score[,c(2,7,10,8,11,9,12)], 10))
```


```{r, fig.width=8}
ggplot(CSZ_score, aes(x=z_score_total)) + geom_histogram(binwidth=.5, colour="black", alpha=.3, fill="blue") +
       geom_vline(data=CSZ_score, aes(xintercept=mean(z_score_total)), linetype="dashed", size=1, colour="red") +
       ggtitle("Z-score distribution of students' total scores is slightly right skewed") +
       labs(x="Total Score (Z-score)",y="Number of Students")
```


####Does poverty affect how well a student does on tests?  
```{r, fig.width=8}
# Scatter Plot: Z-Score of Total by Class Size and Free Lunch
ggplot(CSZ_score, aes(x=z_score_total,fill=free_lunch)) + geom_histogram(binwidth=.5, colour="black", alpha=.3, position="identity") +
       scale_fill_manual(values=c("red", "green")) +
       geom_vline(data=CSZ_score, aes(xintercept=mean(z_score_total)), linetype="dashed", size=1, colour="red") +
       ggtitle("Student participating in free lunch programs severly underperform") +
       labs(x="Total Score (Z-score)",y="Number of Students", fill="Free Lunch") + 
       theme(legend.position="bottom")
```


#Association Rule Mining

####Preparing the data for Apriori Modeling
```{r}
CSApriori <-  CSdata[,c(-1,-7,-8)]
CSApriori$total_score <- round(scale(CSApriori$total_score, center = TRUE, scale = TRUE),2)

# Discretize dataset for Teacher Experience
CSApriori$teacher_experience <- cut(CSApriori$teacher_experience, 
                                    breaks = c(-1,0,4,9,14,Inf),
                                    labels=c("0 Years", "1-4 Years", "5-9 Years","10-14 Years","15+ Years"))



# Discretize by Total Score based on Z-Score
CSApriori$total_score <- cut(CSApriori$total_score,
                             breaks = c(-4,-1.51,-0.51,0.5,1.50,Inf), 
                             labels=c( "Poor","Below Average", "Average", "Above Average","Excellent"))

CSApriori_Summary <-  summary(CSApriori)
colnames(CSApriori_Summary) <- c("Class Size", "Student Gender", "Free Lunch", "Race", "Teacher Experience", "Total Z-Score")
kable(CSApriori_Summary)

```


####Structure of Apriori Dataset 
```{r}
str(CSApriori)
```


####Load the transformed data into the apriori algorithm
```{r}
rules = apriori(CSApriori, parameter = list(supp = 0.001, conf = 0.9, maxlen = 3))
```


####Get summary info about all rules
```{r}
summary(rules)
```

####Sorting rules by highest confidence and lift (Top 5 rules)
```{r}
rules<-sort(rules, by=c("confidence", "lift"), decreasing=TRUE)
inspect(rules[1:5])
```

####Frequently occuring items with "Excellent Test Scores":
```{r}
rules<-suppressWarnings(apriori(data=CSApriori, parameter=list(supp=0.001,conf = 0.08, maxlen=3), 
               appearance = list(default="lhs",rhs="total_score=Excellent"),
               control = list(verbose=F)))
rules<-sort(rules, decreasing=TRUE,by=c("confidence", "lift"))
inspect(rules[1:5])
```


####Frequently occuring items with "Above Average Scores":
```{r}
rules<-suppressWarnings(apriori(data=CSApriori, parameter=list(supp=0.001,conf = 0.08, maxlen=3), 
               appearance = list(default="lhs",rhs="total_score=Above Average"),
               control = list(verbose=F)))
rules<-sort(rules, decreasing=TRUE,by=c("confidence", "lift"))
inspect(rules[1:5])
```

####Frequently occuring items with "Average Scores":
```{r}
rules<-suppressWarnings(apriori(data=CSApriori, parameter=list(supp=0.001,conf = 0.08, maxlen=3), 
               appearance = list(default="lhs",rhs="total_score=Average"),
               control = list(verbose=F)))
rules<-sort(rules, decreasing=TRUE,by=c("confidence", "lift"))
inspect(rules[1:5])
```


####Frequently occuring items with "Below Average Scores":
```{r}
rules<-suppressWarnings(apriori(data=CSApriori, parameter=list(supp=0.001,conf = 0.08, maxlen=3), 
               appearance = list(default="lhs",rhs="total_score=Below Average"),
               control = list(verbose=F)))
rules<-sort(rules, decreasing=TRUE,by=c("confidence", "lift"))
inspect(rules[1:5])
```


####Frequently occuring items with "Poor Scores":
```{r}
rules<-suppressWarnings(apriori(data=CSApriori, parameter=list(supp=0.001,conf = 0.08, maxlen=3), 
               appearance = list(default="lhs",rhs="total_score=Poor"),
               control = list(verbose=F)))
rules<-sort(rules, decreasing=TRUE,by=c("confidence", "lift"))
inspect(rules[1:5])
```


#Decision Tree

####Load the Dataset
```{r}
options(stringsAsFactors = FALSE)

CS_dt <- CSdata[,-c(1,7,8)]
kable(head(CS_dt))

```


####Build and tune decision tree models
```{r}
fit <- rpart(class_size ~ total_score,
             data=CS_dt,
             method="class")
summary(fit)
```

####Testing the Accuracy of Data Actual vs. Predicted
```{r}
predicted= predict(fit,CS_dt, type ="class")
submit <- data.frame(class_size = CS_dt$class_size, Predicted_size = predicted)
confusionMatrix(predicted, factor(CS_dt$class_size))
```

####Visualize the Decision Tree
```{r, fig.width=8}
fancyRpartPlot(fit)
```



#k-Means Clustering

####Preparing the Dataset for Clustering
```{r}
clust_data <- CSdata[,-1]

#Turning nominal data into binary
clust_data$gender <- ifelse(clust_data$student_gender=='boy',1,0)
clust_data$lunch <- ifelse(clust_data$free_lunch=='yes',1,0)
clust_data$white <- ifelse(clust_data$race=='white',1,0)
clust_data$black <- ifelse(clust_data$race=='black',1,0)
clust_data$other <- ifelse(clust_data$race=='other',1,0)

#Removing columns with nominal data
clust_data <- clust_data[,-c(2:4,6:7)]

#standartizing all numeric data as Z-Score
clust_data$teacher_experience <- scale(clust_data$teacher_experience, center = TRUE, scale = TRUE)
clust_data$total_score <- scale(clust_data$total_score, center = TRUE, scale = TRUE)
clust_data$gender <- scale(clust_data$gender)
clust_data$lunch <- scale(clust_data$lunch)
clust_data$white <- scale(clust_data$white)
clust_data$black <- scale(clust_data$black)
clust_data$other <- scale(clust_data$other)

kable(head(clust_data, n= 6))

```

####Removing Class Size Field
```{r}
clust_dataonly <- clust_data[,-1]
kable(head(clust_dataonly, n= 6))
```

####Clustering Model where k=3
```{r}
k = 3
set.seed(4)
kmeansModel <- kmeans(clust_dataonly, k,iter.max = 500)
```

####Cluster Sizes
```{r}
kmeansModel$size
```

####Cluster Centers
```{r}
kmeansModel$centers
```

####Get the clusters that were assigned to each group
```{r}
clusterGroups <- data.frame(clust_data,kmeansModel$cluster)
kable(head(clusterGroups))
```


####Aggregated Cluster Data by Class Size
```{r}
clust_agg <- aggregate(clusterGroups[1], by=list(clusterGroups$kmeansModel.cluster,clusterGroups$class_size), FUN=length)
colnames(clust_agg) <- c("Cluster", "Class Size", "Count")
kable(clust_agg)
```


####Cluster Plot
```{r, fig.width=8}
clusplot(clust_dataonly, kmeansModel$cluster,color=TRUE,
                shade=TRUE, labels=4, lines=0 )
```












